Sodium silicate and hexametaphosphate promote the release of (oxyhydr)oxide nanoparticles from corroding iron

### 2022-09-20 Rmarkdown/paper.Rmd
resubmission to es-adv

Method parameters are summarized in Table \@ref(tab:table-run-params). The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward, it was set to decay linearly over 2 minutes to 0.1 mL min^-1^. The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; then it was set to zero for the rinse step (the final 10 minutes of the run).

@@

Sodium silicate and hexametaphosphate suspensions yielded similar iron fractograms (Figure \@ref(fig:figure-fff)), with primary and secondary light scattering peaks. These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes. Iron was also present in a broad peak, eluting between approximately 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm. This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017]. In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at approximately `r unique(round(peak_pos_p$time))` minutes, which aligns with the peak retention of sodium hexametaphosphate in pure water (Figure S3).

The control suspension---prepared without either sodium silicate or hexametaphosphate---did not contain iron colloids that were detectable by FFF-ICP-MS (Figure \@ref(fig:figure-fff)). This is consistent with previous research showing that, in the absence of a stabilizing agent---polyphosphate, sodium silicate, natural organic matter, etc...---iron oxide nanoparticles tend to form aggregates with diameters greater than 450 nm, the upper size limit considered here [@zhang_impact_2009; @mylon_influence_2004; @rushing_effects_2003].

@@

While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by hexametaphosphate may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions. And since recovery of P from a sodium hexametaphosphate standard was 111%, the partial P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1). The partial recovery of iron may be due to chelation by trimetaphosphate [@rasmussen_phosphate_1986], which would also escape retention by the membrane. In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].

@@

Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively. The strong apparent relationship between iron and lead is consistent with much previous research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010]. Given that iron colloids with diameters smaller than 450 nm were not detectable in the control suspension, it was excluded from the partitioning experiment. The larger aggregates it contained, though, would also be expected to adsorb lead [@mckenzie_adsorption_1980].

@@

```{r figure-pb, fig.cap="Fractograms representing duplicate sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^. Light scattering and absorbance fractograms are displayed with the raw detector responses on the y-axis.", fig.height=4}

@@

Moreover, even sequestrants that do not form strong complexes with lead (e.g., sodium silicates) may increase lead release. In a recent model distribution system study [@li_impact_2020], for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream. In the long term, silicate treatment may decrease iron release even if it initially does the opposite [@rushing_effects_2003], but anticipating and preparing for the short-term implications highlighted here will help protect water quality. Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.

@@

>>>>

Method parameters are summarized in Table \@ref(tab:table-run-params). The crossflow was maintained at 2.0 mL min^-1^ for the first 31 minutes of each run (including the focus and transition periods); afterward, it was set to decay linearly over 2 minutes to 0.1 mL min^-1^. The crossflow was then maintained at 0.1 mL min^-1^ for 10 minutes; afterward it was set to zero for the rinse step (the final 10 minutes of the run).

@@

Sodium silicate and hexametaphosphate suspensions yielded similar iron fractograms (Figure \@ref(fig:figure-fff)), with primary and secondary light scattering peaks. These peaks had maxima, based on ^57^Fe intensity, at `r pks_txt$min_1`--`r pks_txt$max_1` and `r pks_txt$min_2`--`r pks_txt$max_2` minutes. Iron was also present in a broad peak, eluting between 15 and 30 minutes, that did not coincide with a significant light scattering signal but did adsorb UV light at 254 nm. This peak may represent iron nanoparticles that are too small to be quantified using the online MALS detector [@cuss_af4_icpms_2017]. In sodium hexametaphosphate suspensions, the principal ^31^P peak eluted at approximately `r unique(round(peak_pos_p$time))` minutes, which matches the peak retention of sodium hexametaphosphate in pure water (Figure S3).

The control suspension---prepared without sodium silicate or hexametaphosphate---did not contain iron colloids that were detectable by FFF-ICP-MS (Figure \@ref(fig:figure-fff)). This aligns with research showing that, in the absence of a stabilizing agent---a sequestrant, for instance---iron oxide nanoparticles tend to form aggregates with diameters greater than 450 nm, the upper size limit considered here [@zhang_impact_2009; @mylon_influence_2004; @rushing_effects_2003].

@@

While FFF recovered essentially all of the iron in sodium silicate suspensions, complexation by polyphosphate species may have resulted in losses of iron to waste in separations of hexametaphosphate suspensions. And since recovery of ^31^P from a sodium hexametaphosphate standard was 111%, the partial ^31^P recovery from the iron oxide suspension suggests that hexametaphosphate hydrolyzed to orthophosphate or depolymerized to trimetaphosphate [@bell_hydrolysis_1947], species that readily pass through the FFF membrane to waste (Figure S1). The partial recovery of iron may be due to chelation by trimetaphosphate in particular [@rasmussen_phosphate_1986]: the resulting complex would also escape retention by the FFF membrane. In drinking water systems, hexametaphosphate reversion is expected to be significant, especially in older pipe networks [@holm_metaphosphate_2003].

@@

Lead partitioned readily to suspended iron oxide colloids when added to the sodium silicate and hexametaphosphate suspensions (Figure \@ref(fig:figure-pb)), with variation in ^57^Fe explaining r^2^ = 99 and 93% of variation in ^208^Pb, respectively. The strong apparent relationship between iron and lead is consistent with research documenting lead adsorption to iron (oxyhydr)oxides [@mckenzie_adsorption_1980; @ainsworth_cobalt_1994] and, more specifically, elevated lead in iron-rich drinking water [@li_impact_2020; @trueman_galvanic_2017; @trueman_understanding_2016; @masters_increased_2015; @mcfadden_contributions_2011; @deshommes_source_2010]. Given that iron colloids with diameters smaller than 450 nm were not detectable in the control suspension, it was excluded from the partitioning experiment. The larger aggregates it contained, though, would also be expected to adsorb lead [@mckenzie_adsorption_1980].

@@

```{r figure-pb, fig.cap="Fractograms representing sodium silicate and hexametaphosphate suspensions; each is paired with an equivalent suspension dosed with 200 µg Pb L^-1^. Light scattering and absorbance fractograms are displayed with the raw detector responses on the y-axis.", fig.height=4}

@@

Moreover, even sequestrants that do not form strong complexes with lead may increase lead release. In a recent model distribution system study [@li_impact_2020], for instance, an abrupt increase in the dose of sodium silicate was followed by elevated iron release from corroded cast iron pipe sections and lead release from lead pipes downstream. In the long term, silicate treatment may decrease iron release even if it initially does the opposite [@rushing_effects_2003], but anticipating the short-term effects highlighted here will help protect water quality. Testing that does not account for interactions between lead and iron, then, may underestimate the true risk of elevated lead associated with sequestration.

@@

# Author contributions

Benjamin Trueman: investigation, data curation, formal analysis, visualization, writing--original draft; Javier Locsin: investigation, methodology, writing--review & editing; Evelyne Dor&eacute;: investigation, methodology, writing--review & editing; Kalli Hood: investigation, methodology, writing--review & editing; Graham Gagnon: funding acquisition, writing--review & editing.

# Conflicts of interest

There are no conflicts to declare.
